---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

```{r global_options,echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, results="asis", warning=FALSE)
```

## Loading and preprocessing the data
_Note: The csv file was already extracted from the activity.zip file stored in the forked github branch_

Extracting the data from activity.csv

```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
#activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Histogram of the total number of steps each day, ignoring NA values.
Intervals are blocks of ~2000.

```{r}
hist(with(activity, tapply(steps, date, sum)), 
     main="Activity Monitoring Step Counts", xlab="Steps (per day)", ylab="Frequency",
     breaks = 10)
```

Mean and median steps per day

```{r}
library(xtable)
library(dplyr)
steps_by_date <- activity %>% group_by(date) %>% summarise(mean=mean(steps), median=median(steps, na.rm=TRUE))
print(xtable(steps_by_date), type="html")
```

## What is the average daily activity pattern?

Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

8:35 is the 15-minute time interval with the highest number of steps on average across all days

```{r daily_pattern, results="markup"}
avg_steps_by_interval <- activity %>% group_by(interval) %>% summarise(mean=mean(steps, na.rm=TRUE))
plot(avg_steps_by_interval$interval, avg_steps_by_interval$mean, 
     type="l", col="blue", main="Average Steps by Time Interval", xlab="Interval", ylab="Average Steps")

with(avg_steps_by_interval, subset(avg_steps_by_interval, mean==max(mean)))
```

## Imputing missing values

There are 2304 missing values for steps in the activity dataset
```{r missing1, results="markup"}
summary(activity$steps)
```

Replace the NAs with the median for each day.  
Recall from the table above, all medians are 0.  

```{r}
activity_cleaned <- activity
activity_cleaned[is.na(activity_cleaned)] <- 0
```

Revisiting the histogram and mean/median calculations.
You can see that the mean/median values are the same where there was previously data.
So no additional steps have been the result of imputing the data.

```{r}
steps_by_date_cleaned <- activity_cleaned %>% group_by(date) %>% 
        summarise(mean=mean(steps), median=median(steps, na.rm=TRUE))
print(xtable(steps_by_date_cleaned), type="html")
```

```{r daily_pattern_cleaned, results="markup"}
avg_steps_by_interval_cleaned <- activity_cleaned %>% group_by(interval) %>% 
        summarise(mean=mean(steps, na.rm=TRUE))
plot(avg_steps_by_interval_cleaned$interval, avg_steps_by_interval_cleaned$mean, 
     type="l", col="blue", main="Average Steps by Time Interval (Cleaned)", xlab="Interval", ylab="Average Steps")
```


## Are there differences in activity patterns between weekdays and weekends?

Prepare the weekday and weekend data sets

```{r}
activity_cleaned$date <- as.Date(activity_cleaned$date)
activity_cleaned$weekday<-as.numeric(format(as.Date(activity_cleaned$date),"%w"))
activity_cleaned$weekday<-factor(activity_cleaned$weekday <= 5, 
        levels=c("TRUE","FALSE"), labels=c("weekday","weekend"))

interval_by_factor <- activity_cleaned %>% group_by(interval, weekday) %>% 
        summarise(mean=mean(steps, na.rm=TRUE))
```


The difference between the weekday and the weekend is that typically there is less activity after the early morning when compared to weekends.

```{r weekday_weekend, results="markup"}
library(lattice)
xyplot(mean ~ interval | weekday, data=interval_by_factor, layout=c(1,2), type="l")
```
